---
title: Exercices autour de data.table
author: ~
date: '2017-10-23'
slug: data.table
categories: ["R"]
tags: ["workshop", "package", "tidy", "data.table", "dplyr"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

L'atelier du jour avait pour objectif la découvert du package [data.table](https://cran.r-project.org/web/packages/data.table/index.html) au travers d'[exercises proposés lors de la dernière édition de UseR](https://user2017.sched.com/event/Aw0L/datatable-for-beginners). Les slides avec les énoncés sont [disponibles ici](https://www.google.fr/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=0ahUKEwiZv7DE6I3XAhWCuBoKHcN5CREQFggrMAA&url=https%3A%2F%2Fchannel9.msdn.com%2FEvents%2FuseR-international-R-User-conferences%2FuseR-International-R-User-2017-Conference%2Fdatatable-for-beginners&usg=AOvVaw3N-q_5oW0-yMm8uz9hVisu).

L'apport de notre atelier est d'avoir systématiquement  essayé de traiter les questions avec 
  - les fonctionalités `R` de *base*
  - les fonctionalités offertes par le *tidyverse* (principalement *dplyr*)
  - les fonctionnalités du package *data.table*
  
Pour les animateurs : vous êtes libre de compléter le post de blog. Nous ajouteroins des comparatifs en terme de temps de calcul plus tard (peut-être).

### Préliminaires

C'est parti, on charge oles packages requis

```{r, message=FALSE}
library(data.table)
library(tidyverse)
```

Les exercices seront traités à l'aide de deux tables jouets, dont on propose des contreparties en
  - object `data.frame`
  - object `tibble`
  - object `data.table`
  
```{r}
set.seed(20170703L)
DF1 = data.frame(id = sample(1:2, 9L, TRUE), 
                 code = sample(letters[1:3], 9, TRUE), 
                 valA = 1:9, valB = 10:18, 
                 stringsAsFactors = FALSE)
DF2 = data.frame(id = c(3L, 1L, 1L, 2L, 3L), 
                 code = c("b", "a", "c", "c", "d"), 
                 mul = 5:1, stringsAsFactors = FALSE)
## corresponding data tibble
TB1 <- as.tibble(DF1)
TB2 <- as.tibble(DF2)
## corresponding data tables
DT1 <- as.data.table(DF1)
DT2 <- as.data.table(DF2)

```


### Question 1

_Subset all rows where id column equals 1 & code column is not equal to "c"_

#### base

```{r}
base::subset(DF1, id == 1 & code != "c")
with(DF1, DF1[id == 1 & code != "c",])
```

#### dplyr

```{r}
filter(TB1, id == 1 & code != "c")
```

#### data.table

```{r}
DT1[id == 1 & code != "c"] 
```

### Question 2

_Select valA and valB columns from DF1_

#### base R

```{r}
DF1[, c("valA", "valB")]
```

#### dplyr

```{r}
select(TB1, valA, valB)
```

#### data.table

```{r}
DT1[ , .(valA,valB)]
```

### Question 3

_Get sum(valA) and sum(valB) for id > 1 as a 1-row, 2-col data.frame_

#### base R

```{r}
colSums(DF1[ DF1$id > 1, c("valA", "valB")])
```

#### dplyr

```{r}
TB1 %>% filter(id > 1) %>% select(valA, valB) %>% summarise_all(sum)
```

#### data.table

```{r}
DT1[id > 1, .(sum(valA), sum(valB))]
```

### Question 4

_Replace valB with valB+1 for all rows where code == "c"_

#### base R

```{r}
DF1$valB[DF1$code=="c"] = DF1$valB[DF1$code=="c"] + 1
DF1
```

#### dplyr

```{r}
TB1 <- TB1 %>% mutate(valB = ifelse(code == "c", valB + 1,valB))
## TB1[TB1$code=="c", ] <- filter(TB1, code=="c") %>% mutate(valB= valB+1) # alternative solution
TB1
```

#### data.table

```{r}
DT1[code == "c", valB := valB + 1L]
DT1
```

### Question 5

_Add a new column valC column with values equal to valB^2 - valA^2_

#### base R

```{r}
DF1 <- transform(DF1, valC = valB^2 - valA^2)
## DF1$valC <- DF1$valB^2 - DF1$valA^2 # alternate solution
DF1
```

#### dplyr

```{r}
TB1 <- mutate(TB1, valC = valB^2 - valA^2)
TB1
```

#### data.table

```{r}
DT1[, valC := valB^2 - valA^2]
DT1
```

### Question 6

_Get sum(valA) and sum(valB) grouped by id and code (i.e., for each unique combination of id,code)_

#### base

```{r}
aggregate(.~ id + code, DF1, sum)
aggregate(DF1[, c("valA", "valB")], list(DF1$id, DF1$code), sum)
```

#### dplyr 

```{r}
TB1 %>% group_by(id, code) %>% summarise_all(sum)
```

#### data.table

```{r}
DT1[, .(SumA=sum(valA), SumB = sum(valB)), by=list(id,code) ] 
```

### Question 7

_Get sum(valA) and sum(valB) grouped by id for id >= 2 & code %in% c("a", "c")_

#### base

```{r}
aggregate(.~ id , subset(DF1, id >=2 & code %in% c("a","c"), -code), sum)
```

#### dplyr

```{r}
TB1 %>% 
  group_by(id) %>% 
  filter(id >=2, code %in% c("a", "c")) %>% 
  select(-code) %>% 
  summarise_all(sum)
```

#### data.table

```{r}
DT1[ (id>=2) & code %in% c("a", "c") , .(SumA=sum(valA), SumB = sum(valB)), by=id ] 
```

### Question 8

_Replace valA with max(valA)-min(valA) grouped by code_

#### base

```{r}
DF1 <- transform(DF1, valA = rep(tapply(valA, code, function(x) diff(range(x))), table(code)))
DF1
```

#### dplyr

```{r}
TB1 <- TB1 %>% group_by(code) %>% mutate(valA= max(valA)-min(valA))
TB1
```

#### data.table

```{r}
DT1[, "valA" := max(valA)-min(valA), by=code] 
DT1
```

### Question 9

_Create a new col named valD with max(valB)-min(valA) grouped by code_

#### base

```{r}
DF1 <- transform(DF1, valD = rep(
  by(DF1[, c("valA", "valB")], code, function(x) diff(range(x))), table(code))
)
```

### Question 10

_Subset DF1 by DF2 on id,code column. That is, for each row of DF2$id, DF2$code, get valA and valB cols from DF1. Include rows that have no matches as well._

#### base 

```{r}
merge(DF1, DF2[, c("id", "code")], by = c("id", "code"), all.y=TRUE)
```

### Question 11

_Same as (10), but fetch just the first matching row of DF1 for each row of DF2$id, DF2$code. Exclude non-matching rows._

#### base

```{r}
merge(DF1[!duplicated(DF1[, c("id", "code")]), ],
      DF2[, c("id", "code")], by = c("id", "code"), all.y=TRUE)
```

### Question 12

_For every row of DF2$id, DF2$code that matches with DF1’s, update  valA with valA*mul._

#### base 

```{r}
transform(merge(DF1, DF2, by = c("id", "code"), all.y=TRUE), valA = valA * mul)
```

### Question 13

_Add a new column val to DF1 with values from DF2$mul where DF2$id, DF2$code matches with DF1’s. Rows that don’t match should have NA._

#### base

```{r}
transform(merge(DF1, DF2, by = c("id", "code"), all.x = TRUE, sort = FALSE), mul = mul)
```
